\section{Variational Queries for Variational Case Studies}
\label{app:q}

%\point{For each database, we provide a set of \emph{variational queries}.}\\
%- describe what you mean by variational query\\ --> background
%- describe syntax or formally define it\\ --> background
%- mention that we provide queries in VRA too and also encodings of them in Haskell

Variation in software affects not only databases but also how developers and
database administrators interact with databases.
%
Since different software variants have different information needs, developers
must often write and maintain different queries for different software
variants. Moreover, even if a particular information need is similar across
variants, different variants of a query may need to be created and maintained
to account for structural differences in the schema for each variant.
%
Creating and maintaining different queries for each variant is tedious and
error-prone, and potentially even intractable for large and open-ended
configuration spaces, such as most open-source projects~\cite{ATW18poly}.


In this section, we illustrate how variation in software leads to variation in
information needs and variation in the queries that realize those information
needs. We also show how variational information needs can be captured by
\emph{v-queries} written in VRA.
%expressive query languages that account
%for variation explicitly and link variation in software and databases to
%queries by using the same feature names and configuration space.
%
For each case study, we provide a set of v-queries.
% that realize the
%information needs of the different variants of the corresponding software
%systems.
%
We present only a  sample of the queries here for space reasons,
yet we provide the full query sets in the GitHub repository linked in \secref{intro}.
%online and will be linked to throughout this
%section.
%
%The full query sets capture all of the information needs described in the
%papers that  we base our variation scenarios on. It is important to note that
%this makes our query sets potentially biased toward queries containing more
%variation points since the focus of the papers is on variational parts of the
%system (feature interactions in the case of \citet{Hall05} and changes over
%time for \citet{prima08Moon}). A complete query set, capturing \emph{all}
%information needs for each scenario might contain more plain queries, that is,
%queries that perform the same way over all variants.
%%
%However, we do not believe this bias is harmful for the role the case studies
%are intended to serve, namely, motivating and evaluating variational database
%systems. For this role, queries that contain variation are more useful than
%plain queries, and additional plain queries can likely be more easily generated
%if needed.


We distribute the v-queries in two formats: (1) VRA, introduced in \secref{background}, 
encoded in the format used by
our VDBMS tool, and (2) plain SQL queries with embedded \cpp{ifdef}
annotations to capture variation points.
%
The SQL format provides queries for studying variational data 
independently of VDBMS tool,
%will be more immediately useful for other researchers
%studying variational data independently of our VDBMS tool, 
but we use VRA in
this proposal for its brevity.
%because it is much more concise.

\input{sections/enronqs}
\input{sections/empqs}
%\input{sections/qprops}
