\subsubsection{Variational Schema}
\label{sec:vsch}


A variational schema captures variation in the structure of a database by indicating which attributes and relations are included or excluded in which variants.
%
%\point{using annotated elements to show variability in schema.}
%Variation can exist in the structure of data, i.e., the schema.
%As motivated in \secref{mot}, schema variations include/exclude relations/attributes. 
To achieve this we annotate attributes, relations, and the schema itself with feature expressions,
which describe the condition under which they are present.
%
A \emph{\vrelTxt\ schema (v-relation schema)}, \vRelSch, is a relation name
accompanied with an annotated variational set of attributes:
%\centerline{
$\synDef {\vRelSch} \vRelSchSet \eqq \vRelDef$.
%}
The presence condition of the v-relation schema, \dimMeta, determines the
set of all possible relation schema variants for relation \vRel.
%A \emph{variational attribute set}, \vAttList, is a variational set of attributes.
%%i.e.,
%%$\synDef \vAttList \vAttSet \eqq \optAtt, \vAttList \myOR \empAtt$,
%%where \empAtt\ denotes an empty attribute. 
%%\exref{vsch} illustrates creating
%%a v-schema.
%
A \emph{\vschTxt\ (v-schema)} is an annotated set of v-relation 
schemas:
%\centerline{ 
$\synDef \vSch \vSchSet \eqq \vSchDef$.
%} 
The presence condition of the v-schema, \fModel, determines all  
configurations for non-empty schema variants. We call such configurations
\emph{valid} configurations. The v-schema's presence condition 
is the VDB feature model since it captures the relationship
between features of the underlying application and their constraints,
%as explained in \secref{encode-var} and introduced as the VDB feature model. 
Hence, the v-schema defines all valid schema variants of a VDB. 


\begin{example}
\label{eg:vsch}
$\vSch_1$ is the v-schema of a VDB including only relations \empacct\ and \ecourse\ in the last two rows
of \tabref{mot}, where only features are \vFour, \vFive, \edu, \tFour, \tFive.
Note that attributes that exist conditionally are annotated with a feature expression
to account for such a condition, e.g., the \salary\ attribute only exists when \vFive\ = \t.
%
\begin{align*}
\vSch_1 &=
\{ \empacct ( \empno, \hiredate, \titleatt, \deptno, \annot [\vFive] \salary, 
%&\hspace{55pt}
 \annot [\edu] \isstudent,
\annot [\edu] \isteacher )^{\vFour \vee \vFive}\\
%
&\hspace{17pt} ,\ecourse ( \cno, \cname, \annot [\tFive] \deptno )^{\tFour \vee \tFive} \}^{\fModel_1}\\
\fModel_1 &= 
\paran {\neg \edu \wedge \paran {\vFour \oplus \vFive}
 }
\vee
\paran {\edu \wedge \paran {\vFour \oplus \vFive}
\wedge
\paran {\tFour \oplus \tFive}
}
\end{align*}
where \ensuremath{\fModel_1} allows only one temporal feature for each schema column 
be enabled at a given time.
\end{example}

%\input{sections/motVsch}

%\textbf{Hierarchal structure of feature expressions in a v-schema:}
The presence of an attribute naturally depends on the presence of its parent v-relation, which in turn depends on the feature model.
% presence of the v-schema. 
%
Thus, the presence condition of an attribute is  
the conjunction of its presence condition with its v-relation's presence condition
and the feature model.
%
Similarly,
the presence condition of a v-relation is the conjunction of its
presence condition and the feature model.
%
In other words, the annotated attribute \optAtt\ of v-relation \vRel\ with 
$\dimMeta_r = \getPC \vRel$
defined in the v-schema \vSch\ with feature model \fModel\
is valid if: $\sat {\dimMeta \wedge \dimMeta_r \wedge \fModel }$.
For example, the \isstudent\ attribute described in \exref{vsch} 
is only valid if its presence condition is
satisfiable, i.e.,
\ensuremath {\sat {\edu \wedge (\vFour \vee \vFive) \wedge \fModel }}.

%\subsubsection{Configuring a V-Schema}
%\label{sec:conf-vsch}
In essence, a v-schema is a systematic  
compact representation of all schema variants of the 
underlying application of interest.
%, e.g., SPL, that encodes 
%the variation effectively inside the database schema by means of 
%feature expressions.
%Consequently, v-schemas relieve the
%need to define an intermediate schema and state mappings 
%between it and source schemas, like the approach that 
%data integration systems employ. 
%One can still obtain the 
A specific pure relational schema for
a database variant can be obtained 
by \emph{configuring} the v-schema with that variant's configuration.
We define the configuration function for v-schemas and its elements in \figref{vdb-conf}.
For example, consider the v-schema in \exref{vsch}.
Configuring the variational attribute set of the \empacct\ v-relation for 
the variant \setDef {\vFive}, i.e., 
\ensuremath {\olSem [\setDef {\vFive}] {\empno, \hiredate, \titleatt, \deptno}},
yields the 
%relational 
attribute set of
\ensuremath {\setDef {\underline \empno,\underline \hiredate,\underline \titleatt,\underline \deptno }}.



