First and foremost, I am very thankful to my advisor Eric Walkingshaw for 
piquing my interest in programming languages through our early days of collaboration
and functional programming by suggesting that I implement my system in Haskell.
%
Eric's insistence on exploring my ideas  and
experimenting with them--even when he knew I am heading 
to the wrong path--allowed me to become more independent in my research over time. 
%
His advice and feedback on research and scientific writing taught me lots of 
valuable lessons. His dedication 
to making interdisciplinary programming languages research accessible
to and understandable by a wide audience through meticulous editing of papers
and presentations has been nothing short of exemplary for me.
%
His support and encouragement to tickle my curiosity by taking classes outside of
OSU and exploring various resources in our reading group allowed me to find the beautiful corner
of programming languages that I absolutely enjoy.
%
His care for my emotional and mental well-being significantly improved the quality 
of my life throughout my Ph.D. in OSU. 
%
Finally, his dedicated, innovative, and caring research and teaching style 
had a significant role in forming the researcher and (sometimes) the teacher I am today. 
%
I am forever grateful for his support, guidance, enthusiasm, and encouragement through 
the years.
%Thank you for believing in me even at the 

%
A very special thank you to my partner Jeffrey M. Young for his love, support, and encouragement. 
%
Thank you for your technical help--from me learning Haskell when I initially joined the PL group to plotting the diagrams in this thesis.
Thank you for discussing research ideas and problems,
practicing many presentations, and studying various books and papers with me. 
Your dedication to learning, helping others, and community has actively made the PL group at OSU a better place.
%Thank you for your support through emotionally tough times.
%, from dealing with racist policies
%to depression from a pandemic and a harsh immigration process. 
%I am proud of us for finishing our PhDs together during a pandemic, OSU's lack of support of our group, and immigration process. 
Thank you for an amazing graduation gift. I look forward to dropping the Ph.D. weight! 
%And congratulation for graduating your PhD! 
%I appreciate our graduate school years, for getting through it together during a pandemic and dealing with consequences of university's lack of support for the PL group due to partly 
%not  understanding the importance of the field. 
Finally, thank you for sharing your life with me. You make my life sweeter every day.
I look forward to our life together, out of the brutality of graduate school. 


I would like to express my gratitude to the many OSU faculty members and peers during my time here. 
A special thank you to Glencora Borradaile for her unwavering support 
of graduate students. Also, a very special thank you to Martin Erwig for many 
constructive advice and discussions on 
research in either classes or meetings. His dedication to research and teaching 
greatly influenced the researcher in me.
Thank you to the members of my program committee--Martin Erwig, 
Alan Fern, Amir Nayyeri--for all their support and advice. 
%
Thank you to collaborators for professional help and feedback including
Arash Termehchy, Qiaoran Li, and Fariba Khan. A very special thank you
to my colleagues in the Lambda reading group and the functional programming club 
for their passion and interesting discussion.
%Their ideas, hard work, and feedback shaped the research and developments presented
%in this thesis

%
Thank you to other mentors. 
Thank you to David Christiansen for a great dependent typed languages class at PSU
which triggered my enthusiasm to pursue dependent types and formal verification further. It definitely
worth the drive!
And thank you to Jose Calderson for honest open conversations on building a research career. He is
a breathe of fresh air in today's humanity.
%


Thank you to my strong amazing ladies--Lily Ranjbar, 
Nasim Adami, Mahtab Aboufazeli, and Saereh Mirzababaei--for their constant kindness, 
support, and advice during my Ph.D. years. Thank you for always being there for me 
through the good and the bad. Life has thrown some harsh obstacles my way and I 
would not have been able to overcome them without each and every one of you. 
%
%Thank you Lily and Ali Jafarnejad for your continuous support through the years--from 
%hosting me when I came to the US to throwing me a wedding during a pandemic. 
%%
%Thank you Nasim 
%and Ali Mohtat for your many wise words and kind support, especially when life --from
%holding my hand when I was tired 
%of putting up with the unfair single-entrance Visa and was on the edge of leaving the US to 
%present my work in Europe to translating my wedding for my parents.
%%
%Thank you Mahtab for simply making my life sweeter, both figuratively and literally. 
%%
%Thank you Saereh for being a role model of a strong independent Persian woman. You are inspiring.
%% 
%%Thank you Reyhaneh for many conversations, either checking up on me or philosophizing about life.

%
I would like to thank my friends for 
check-ins, many conversations, professional and personal help, sharing lots of coffee (sometimes over complaining about grad school), hangouts, and movie nights.
Thank you Ali Mohtat, Ali Jafarnejad, 
Arezoo Rajabi, Reyhaneh Shahmohamadi, Wesley Alexis,
Beatrice Moissinac, Mandana Hamidi-Haines, Ralph Haines, 
Esther and Hamid Mahmoodabadi, Hadi Hashemi, Ben McCamish, Vahid Ghadakchi, Reza Ghaeini, Farzad Zafarani, Mike McGirr, Ameya Ketkar, Colin Shea-Blymyer, and Will Maxwell.
%
%Thank you to my family all over the world for their support, hospitality, and care.
%A special thank you to my sister for her love and care. 

I would also like to thank  the CrossFit 97333 community
and its coaches. 
%including Derek, Austin, Michael Hernandez, Tanna, Sean, Cassie, and Caitlin.
%Katherine Eastlake, Jeremey \TODO{blah}, and David and Heidi Evaland.
Thank you for providing an environment that 
allowed me to escape the Ph.D. life even just for a couple of hours a week. Thank you
for pushing me physically and mentally through lots of tough, yet, enjoyable workouts. 
%And thank you
%for being such inspirational individuals in and out of the box.
%I am very excited to get after it again.
% I have made in Corvallis. 
%
%Thank you Wesley \TODO{blah} for deep conversations and movie nights.
%%
%Thank you Vahid Ghadakchi for your feedback.
%%
%Thank you Ben McCamish for many coffee hours during which we got lots of grad school complains off our chests.
%%
%Thank you Beatrice \TODO{blah} for your advice, support, and check-ins. 
%%
%Thank you Mandana and Ralph Haines for introducing me to crossfit, helping me with immigration
%paperwork, and kind humane conversations. 
%%
%Thank you Esther and Hammid \TODO{blah} for many dinners and hangouts that took my mind off of
%graduate school and for helping me with immigration paperwork.
%%
%Thank you to Reza Ghaeini and Farzad Zafarani for help with coursework during my first term here. 
%
%including Maryam Soltanolkotabi, Hamed Asghari, Mahdi Soltanolkotabi, Amirreza Ataei, 
%Mohammad Reza Bazrafshan, Mehri Ataei, Reza Khodadadi, Tahereh Bazrafshan, 
%Mahmood Soltanolkotabi, and my loving sister Parvin Ataei. 

Last but not least, I am eternally grateful 
to my parents directing me into this path and prioritizing my education over the years. 
Thank you for giving me a prospering life and the freedom to grow in the path I desired.
I appreciate all your sacrifices and I am deeply sorry for the pain of separation over the last six 
years. A special thank you to my sister for her love and care. I cannot wait to be reunited with all of you.


\begin{comment}
I am very grateful to my advisor Emmanuel Cand`es for piquing my interest in applied
math and statistics through his papers as an undergrad and later on throughout my
Ph.D. at Stanford. Emmanuel?s insistence on tackling difficult yet relevant problems
was ideal for me. His advice and feedback on research and scientific writing-even when
I disagreed with them-were invaluable as it would often force me into the thinking
position. I am also thankful to him for giving me the freedom to pursue eclectic and
sometimes unorthodox research directions which helped me find an independent voice
in research. Finally, Emmanuel?s dedication to making theoretical research accessible
to and understandable by a wide audience through meticulous editing of papers,
presentations and class lectures has been nothing short of exemplary for me.

I would like to express my gratitude to Stephen Boyd, Trevor Hastie, Andrea
Montanari, and David Tse for serving on my thesis committee, as well as many other
professors whom I took classes with at Stanford. I would like to thank Andrea Montanari for many enthusiastic presentations, lectures and classes at Stanford. His depth
and diversity of knowledge made his research presentations/classes a very enriching
and rewarding experience. I am grateful to Trevor Hastie for teaching me statistical
learning through his classes and his fantastic book as well as his support during the
past year. I would also like to thank Stephen Boyd for giving me the opportunity to
teach his class during my second year as a grad student which helped me get a better
grasp of convex optimization early on.

I have been fortunate to have had great peers throughout my education. Countless
peers and friends during undergrad at Sharif EE who with their uniform excellence
kept me motivated and focused on my studies. Special thanks goes to Saber Saleh
Kaleybar and Adel Javanmard my collaborators through many course assignments
and projects. I would also like to thank several fellow graduate students as well as my
group members past and present at Stanford: Carlos Fernandez-Granda, Xiaodong
Li, Adel Javanmard, Yaniv Plan, Mark Davenport, Ewout Van Den Berg, Stephen
Becker, Rina Foygel, Lester Mackey, Veniamin Morgenshtern, Deanna Needell, Reinhard Heckel, Carlos Sing-Long, Vlad Voroninski, Alexandra Chouldechova, Weijie Su,
Kahye Song, Yuxin Chen, and Kenji Nakahira.

Last but not least, I am eternally grateful to my parents and my sister, for their
endless love and support. Without you, I would not be here today.
\end{comment}