\section{Variation in Space: Email SPL Use Case}
\label{sec:enron-vdb}


In our first case study, we focus on variation that occurs in ``space'', that
is, where multiple software variants are developed and maintained in parallel.
In software, variation in space corresponds to a SPL, where many distinct
variants (products) can be produced from a single shared code base by enabling
or disabling features. A variety of representations and tools have been
developed for indicating which code belongs to which feature(s) and supporting
the process of configuring a SPL to obtain a particular variant.


% E: This is repetitive with the intro, but I think it's OK to remind them here
% of what the current bad solutions are. Plus, we have space to fill. :-)

Naturally, different variants of a SPL have different information needs. For
example, an optional feature in the SPL may require a corresponding attribute
or relation in the database that is not needed by the other features in the
SPL.
%
Currently, there is no good solution to managing the varying information needs
of different variants at the level of the database.
%
One possible solution is to manually maintain a separate database schema for
each variant of the SPL. This works for some SPLs where the number of products
is relatively small and the developer has control over the configuration
process. However, it does not scale to open-source SPLs or other scenarios
where the number of products is large and/or configuration is out of the
developer's hands.
%
Another possible solution is to use and maintain a single universal schema that
includes all of the relations and attributes used by any feature in the SPL. In
this solution, every product will use the same database schema regardless of
the features that are enabled. This solves the problem of scaling to large
numbers of products but is dangerous because it means that potentially several
attributes and relations will be unused in any given product. Unused attributes
will typically be populated by \nul\ values, which are a well-known source of
errors in relational databases~\cite{AliceBook}.


VDBs solve the problem by allowing the structure of a relational database to
vary in a synchronous way with the SPL. Attributes and relations may be
annotated by presence conditions to indicate in which feature(s) those
attributes and relations are needed.
%
An implementation of the VDB model might use a universal schema under-the-hood
to realize VDBs on top of a standard relational database management system
(indeed, this is exactly how our prototype VDBMS implementation works), but by
capturing the variation in the schema explicitly, we can validate (potentially
variational) queries against the relevant variants of the variational schema to
statically ensure that no \nul\ values will be referenced.

%Our first use case focuses on variation in space.
The email SPL use case shows the use of VDB to encode the variational
information needs of a database-backed SPL. We consider an email
SPL that has been used in several previous SPL research projects (e.g.\
\cite{apel2013strategies,AlHaj19}).
It develops a variational
schema that captures the information needs of a SPL based on Hall's
decomposition of an email system into its component features~\cite{Hall05}. The
email SPL has been used in several previous SPL research projects (e.g.\
\cite{Apel13:SSP,AlHaj19}). The variational email database is populated using
the Enron email dataset, adapted to fit our variational schema~\cite{Shetty04}.
%
%Our first case study demonstrates the use of a VDB to encode the variational
%information needs of a database-backed SPL.
%
Our use case is formed by systematically combining two pre-existing works:
%
\begin{enumerate}
%
\item 
 We use Hall's decomposition of an email system into its component
features~\cite{Hall05} as high-level specification of a SPL.
%
\item 
 We use the Enron email
dataset\footnote{\url{http://www.ahschulz.de/enron-email-data/}} as 
a source of
a realistic email database.
%
\end{enumerate}
%
In combining these works, we show how variation in space in an email SPL
requires corresponding variation in a supporting database, how we can link the
variation in the software to variation in the database, and how all of these
variants can be encoded in a single VDB.


\subsection{Variation Scenario: An Email SPL}
\label{sec:enron-scenario}


The email SPL consists of the following features from \citet{Hall05}: 
%
\begin{itemize}
%[leftmargin=*]
%\itemsep0em
%
\item 
\addressbook, users can maintain lists of known email addresses with
corresponding aliases, which may be used in place of recipient addresses;
%
\item 
\signature, messages may be digitally signed and verified using
cryptographic keys;
%
\item 
\encryption, messages may be encrypted before sending and decrypted upon
receipt using cryptographic keys;
%
\item 
\autoresponder, users can enable automatically generated email responses
to incoming messages;
%
\item 
\forwardmessages, users can forward all incoming messages automatically to another
address;
%
\item 
\remailmessage, users may send messages anonymously;
%
\item 
\filtermessages, incoming messages can be filtered according to a
provided white list of known sender address suffixes; and
%
\item 
\mailhost, a list of known users is maintained and known users may
retrieve messages on demand while messages sent to unknown users are rejected.
%
\end{itemize}

%\noindent
%
Note that 
Hall's decomposition separates \signature\ and \encryption\ into two
features each (corresponding to signing and verifying, encrypting and
decrypting). Since these pairs of features must always be enabled together
 and
they are so closely conceptually related, 
we 
reduce them to one feature each for simplicity.


The listed features are used in presence conditions within the
variational schema for the email VDB, linking the software variation to
variation in the database.
%
In the email SPL, each feature is optional and independent, resulting in the
  simple feature model $\fModel_\enron = \t$,
% , which is equivalent to \t, 
given as a feature expression.
%%\eric{should we just consider this feature model as true? that is how 
%%we're implementing it.}
%%
%\begin{align*}
%\fModel_\enron
%  &= \t \vee \addressbook \vee \signature \vee \encryption \\
%  &\quad \vee \autoresponder \vee \forwardmessages \\
%  &\quad \vee \remailmessage \vee \filtermessages \vee \mailhost
%\end{align*}
%
The feature model $\fModel_\enron$ is used as the root presence condition of
the variational schema for the email VDB, implicitly applying it to all
relations, attributes, and tuples in the database.

%For the rest of this section, we study a SPL that 
%generates email systems with different features and demonstrate
%how variation appears in its database in addition to how all variants
%of a database can be encoded in a single variational database. We adapt the
%email system explained in~\cite{} by considering the following features:
%...
%
%We consider the simple feature model of:
%...
%The feature model is part of the variational schema too and is applied
%to all its tables and attributes. 

\input{sections/enronvsch}
\input{sections/enronpopulate}
