
\documentclass[onehalf,11pt]{beavtex}
\title{Theory and Implementation of a Variational Database Management System}
\author{Parisa Ataei}
\degree{Doctor of Philosophy}
\doctype{Dissertation}
\department{Electrical Engineering and Computer Science}
\depttype{School}
\depthead{Director}
\major{Computer Science}
\advisor{Eric Walkingshaw}
\submitdate{June ?, 2021}
\commencementyear{2021}
%
\abstract{
Databases used in the same context that are intended to satisfy the same information need
share commonalities while varying in some aspects as well. 
% 
For example, databases used to store information for a software system are intended to 
satisfy the client's information need and they share some parts of the schema, however, 
they still vary in some parts of the schema and most of the content.
%
That is, they are \emph{variants} of a conceptual hypothetical database that captures all 
the \emph{variation} among them.
%
This pattern appears repeatedly in databases. Instances of variation occurring in databases
are: schema evolution, database integration, database versioning, data extraction, and 
software development either using software product lines approaches or not. 
%
While there are specialized approaches to some instances of variation in databases
there are no generic solutions to manage variation in databases.

In this thesis, we answer the question: ``can we abstract out this repeating pattern of 
variation appearing in databases and bring the hypothetical database that captures all
the variation among a number of databases to life?''
%
To this end, we formalize the database that captures all variation of a number of databases,
called a \emph{variational database}, and a query language that allows interaction with
the said database, called \emph{variational relational algebra}. 
\TODO{maybe two languages, another: variational SQL, depends on my timing!}
%
We implement these concepts in \emph{Variational Database Management System (VDBMS)},
demonstrate the feasibility of our concepts by developing two real-world use cases, and 
examine the performance of VDBMS over our two use cases. 
}
%
\acknowledgements{
\TODO{Eric.
Committee.
partner.
abu.
parents.
}
}

\usepackage{algorithm}
\usepackage{algorithmic}

\usepackage{ebproof} %for derivation trees
\usepackage{lscape} %to force a page to be in landscape mode
\usepackage{graphicx}
\usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)
\usepackage{comment}
\usepackage{accents}
\usepackage{lambda,cc}
%\usepackage{mathtools} 
\usepackage{array,multirow}
%\usepackage{hhline}
\usepackage{arydshln}
%\usepackage{color}
%\usepackage[usenames, dvipsnames]{xcolor}
\usepackage{xcolor}
%\usepackage{colortbl}
%\usepackage{booktabs}
\usepackage{mathpartir}
\usepackage{hyperref}
\usepackage[normalem]{ulem}
% commuting diagram packages
\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary{decorations.pathmorphing}
\usepackage{wrapfig}
%\usepackage{blindtext}% for example text here only
\usepackage{enumitem,kantlipsum} %remove enumerate indent
\usepackage{amssymb}%for arrow labels

\usepackage{listings} % for v-sql example code
%
% references
\newcommand{\tabref}[1]{\hyperref[tab:#1]{Table~\ref*{tab:#1}}}
\newcommand{\figref}[1]{\hyperref[fig:#1]{Figure~\ref*{fig:#1}}}
\newcommand{\secref}[1]{\hyperref[sec:#1]{Section~\ref*{sec:#1}}}
\newcommand{\defref}[1]{\hyperref[def:#1]{Definition~\ref*{def:#1}}}
\newcommand{\appref}[1]{\hyperref[app:#1]{Appendix~\ref*{app:#1}}}
\newcommand{\chref}[1]{\hyperref[ch:#1]{Chapter~\ref*{ch:#1}}}
\newcommand{\thmref}[1]{\hyperref[thm:#1]{Theorem~\ref*{thm:#1}}}
\newcommand{\lemref}[1]{\hyperref[lem:#1]{Lemma~\ref*{lem:#1}}}
\newcommand{\exref}[1]{\hyperref[ex:#1]{Example~\ref*{eg:#1}}}
%
%%colors
\definecolor{deepcarminepink}{rgb}{0.94, 0.19, 0.22}%all shared
\definecolor{mediumelectricblue}{rgb}{0.01, 0.31, 0.59}%middlename
\definecolor{frenchblue}{rgb}{0.0, 0.45, 0.73}%lastname
\definecolor{green(munsell)}{rgb}{0.0, 0.66, 0.47}%iceland
\definecolor{violet(ryb)}{rgb}{0.53, 0.0, 0.69}%us,invest
\definecolor{navyblue}{rgb}{0.0, 0.0, 0.5}%v-table
\definecolor{persimmon}{rgb}{0.93, 0.35, 0.0}%iran
\definecolor{Plum}{rgb}{0.78, 0.08, 0.52}%us,bank
\definecolor{ruby}{rgb}{0.88, 0.07, 0.37}%us,bank

\definecolor{light-gray}{gray}{0.95}
\newcommand{\code}[1]{\colorbox{light-gray}{\texttt{#1}}}

%names
\input{names}

%ops
\input{vdbOps}

\begin{document}
\maketitle

\mainmatter


\chapter{Introduction}
\label{ch:introduction}

\section{Motivation and Impact}
\label{sec:motivation}

\section{Contributions and Outline of this Thesis}
\label{sec:contributions}


\chapter{Background}
\label{ch:background}

\section{Relational Database}
\label{sec:rdb}

\section{Formula Choice Calculus}
\label{sec:fcc}

\TODO{what the hell do you want to put in the background?}

\chapter{The Variational Database}
\label{ch:vdb}

\section{Variational Schema}
\label{sec:vschema}

\section{Variational Table}
\label{sec:vtable}

\section{Variational Database}
\label{sec:vdb}

\chapter{Variational Query Language}
\label{sec:vql}

\section{Variational Relational Algebra}
\label{sec:vra}

\section{Variational SQL}
\label{sec:vsql}

\chapter{Variational Database Management System (VDBMS)}
\label{ch:


\chapter{Related Work}
\label{ch:related-work}

\chapter{Conclusion}
\label{ch:conclusion}

%\chapter{Introduction}
%I have done some excellent research \cite{matrix}.
%\section{Introduction to the Introduction}
%\begin{figure}[!ht]
%\centering
%\fbox{\huge Box}
%\caption{Go figure.}
%\end{figure}
%
%\chapter{The Body}
%This is the meat.
%\section{Meat}
%We're born meat and we die meat. Meanwhile, we learn (see Algorithm \ref{alg:learning}).
%
%\begin{algorithm}[h]
%\caption{\textsc{Learning}}
%\label{alg:learning}
%\begin{algorithmic}[1]
%\ENSURE{Optimal policy $\mathcal{C}$}
%	\STATE $\mathcal{C} \gets 42$
%	\RETURN $\mathcal{C}$
%\end{algorithmic}
%\end{algorithm}
%
%\chapter{Conclusion}
%Wow, that really was excellent.
%\section{Fin}
%This is the end, my only friend, the end.


\bibliographystyle{plain}
\bibliography{thesis}

\appendix
\chapter{Proof of Variation-Preserving}
\label{sec:proof}

\end{document}
